Review: Adaptive Introgression Expanded the Genetic Base of Crops during post-Domestication Spread

### 2018-02-11 paper.tex
further edits

The relationship between maize (\emph{Zea mays} ssp. \emph{mays}) and the teosinte \emph{Zea mays} ssp. \emph{mexicana} (hereafter, \emph{mexicana}) offers a prime case study of adaptive wild-to-crop introgression.
Maize was domesticated from \emph{Zea mays} ssp. \emph{parviglumis} (hereafter, \emph{parviglumis})  approximately 9,000 BP in the lowlands of the Balsas River Valley in Mexico \cite{matsuoka2002single}.
From this domestication center, maize spread into the highlands of the Mexican Central Plateau, where it came into sympatry with \emph{mexicana}.
Introgression from \emph{mexicana} to maize in the Central Plateau has been reported based on both morphological \cite {wilkes1977} and molecular \cite{vanHeerwaarden2011, doebley1987} data.
However, Hufford et al. \cite{Hufford2013} first localized \emph{mexicana} introgression to chromosomal regions and provided evidence that it was likely adaptive.
The authors identified nine genomic regions in several maize populations which consistently showed evidence of \emph{mexicana} introgression based on chromosome painting methods including HAPMIX(Figure \ref{fig:introgressionMaize}).
These introgressed segments overlapped with QTLs that had previously been found to control anthocyanin content and leaf macrohairs \cite{lauter2004}, traits known to be adaptive at high elevation.
In a growth chamber experiment, the authors demonstrated that maize populations with \emph{mexicana} introgression showed greater plant height (a proxy for fitness) under highland environmental conditions than populations that lacked introgression.
Height differences were not detected under lowland conditions.

Populations of \emph{mexicana} cannot be found outside of the highlands of Mexico, yet maize has colonized and adapted to high elevation in a number of other regions.
A recent study \cite{wang2017} employed the ABBA-BABA and $\hat{f_{d}}$ statistics to evaluate whether maize with \emph{mexicana} introgression was transferred to other highland regions or whether highland adaptation was obtained \emph{de novo} outside of Mexico.
Overall, analyses revealed that  \emph{mexicana} introgressions were transferred to Meso-American high elevation regions in Guatemala and the southwestern United States, but more distant high elevation regions (\emph{e.g.,} the Andes) showed no \emph{mexicana} ancestry.

Since \emph{mexicana} haplotypes appear to have played an important role historically in adaptation of maize to challenging high-elevation conditions, modern breeding programs may further benefit from drawing on \emph{mexicana} germplasm, particularly in regions like the Andes where \emph{mexicana} alleles are not known to have spread.

\begin{figure}[h]
	\centering
	\includegraphics[width=17.35cm]{./FigsAndTables/Figure1new}
	\caption{Hapmix vectors indicating introgression level from \emph{mexicana} to Mexican highland maize on chromosome 4. Data adopted from Hufford et al. \cite{Hufford2013}. ELP: EL Porvenir; IXT: Ixtlan ; NAB: Nabogame; OPO: Opopeo; PUR: Puruandiro; SAC: Santa Clara; SAP: San Pedro; TEN: Tenango del Aire; XOC: Xochimilco. Insets showed the phenotypes of macrohairs on lowland (red) and highland (blue) maize stems. The black bar indicated one of the QTLs for macrohairs and pigment density in Lauter et al. \cite{lauter2004}.}
	\label{fig:introgressionMaize}
\end{figure}

@@

Barley (\emph{Hordeum vulgare} subsp. \emph{vulgare}) was domesticated at least twice roughly 8,000 to 10,000 BP: once from the wild subsp. \emph{spontaneum} in the Fertile Crescent and once from subsp. \emph{spontaneum} var. \emph{agriocrithon} in Tibet \cite{takahashi1955origin, badr2000origin, azhaguvel2007phylogenetic, haberer2015barley, ren2013tibet, dai2012tibet}.
Presently, the distribution of subsp. \emph{spontaneum} stretches from the eastern Mediterranean through the Middle-East to west-central Asia, spanning clines in temperature, precipitation, soil type, and altitude \cite{nevo2010drought}.
Barley-\emph{spontaneum} hybrids are fertile and commonly found when these taxa co-occur.
Gene flow between wild and domesticated barley is frequent \cite{dai2012tibet}, at times occurring over distances greater than a kilometer \cite{hillman2001new}.

Poets and co-authors \cite{Poets2015} recently investigated the range-wide contribution of wild barley to landraces, assessing both genome-wide and geographical patterns.
This study identified several lines of evidence consistent with wild introgression aiding the dispersal and adaptation of domesticated barley.
Genomic regions of shared ancestry were detected linking particular landraces to numerous wild relative populations, suggesting landraces may have received wild introgression on a continual basis during post-domestication expansion.
However, barley landraces showed an excess of ancestry from nearby wild relatives, indicating a prevalence of local and potentially adaptive gene flow.
Limited linkage disequilibrium and small tracts of identity by state indicated even locally introgressed chromosomal regions are old, perhaps dating to the early expansion of barley  post-domestication.
While these results are suggestive, wild barley haplotypes have yet to be definitely linked to specific local adaptations in landraces.

@@

The details of Asian rice (\emph{Oryza sativa}) domestication are still debated.
Certain genetic and archaeobotanical evidence point toward independent domestications of the two prominent varietal groups \emph{japonica} and \emph{indica} from the wild species \emph{Oryza rufipogon} (\emph{rufipogon} hereafter) in the Yangtze Basin of China 7000 BP and the Indian Ganges plain 4000 BP, respectively \cite{fuller2010consilience}, with a potential third domestication event giving rise to \emph{aus} in Bangladesh or central India \cite{civavn2015three}.
Other studies support a single domestication occurring 8,200-13,500 BP in the Yangtze Basin, with later divergence of \emph{japonica} and \emph{indica} \cite{molina2011molecular, Huang2012}.
Huang and colleagues \cite{Huang2012} developed a genetic map of rice variation, which they used to measure genetic distance between a range-wide sample of wild and domesticated rice, finding \emph{japonica} was domesticated near the Pearl River in Guangxi province, China, and that \emph{indica} was likely the result of hybridization between \emph{japonica} and local \emph{rufipogon} populations in Southern and South-eastern Asia.
In a re-examination of these same data, \cite{civavn2015three} found evidence supporting independent domestications of \emph{japonica}, \emph{indica}, and \emph{aus}, as well as a hybrid origin (\emph{japonica} x \emph{aus}) of \emph{aromatic} rice.
A recent third analysis by Choi and co-authors \cite{choi2018multiple} compared these two disparate results and the methods utilized therein, concluding that domestication alleles (including \emph{LABA1}, \emph{PROG1}, and \emph{sh1}) arose via a single domestication event of \emph{japonica}, and that these alleles introgressed into several wild \emph{rufipogon} subpopulations which thereby became the progenitors of other Asian rice varieties.

The findings of \cite{choi2018multiple} run in parallel with a hypothesis posited by Vaughan and colleagues \cite{vaughan2008evolving}.
According to this hypothesis, domestication alleles arose in a single cultivated rice population and subsequently introgressed into diverse cultivated populations (some \emph{japonica}-like, some \emph{indica}-like) in the region.
As these fully domesticated populations spread into new environments, they potentially received introgression from locally-adapted wild relatives, retaining alleles that improved fitness.

%Despite many conditions favorable to their prevalence, to date, individual examples of adaptive alleles naturally introgressing into domesticated Asian rice are not clear (though \cite{zhao2010genomic} emphasizes the necessity of research on this topic).
%Contemporary populations of wild rices show evidence of this history of local adaptation.
%Wild \emph{rufipogon} and \emph{nivara} both maintain high genotypic and phenotypic diversity and exhibit locally-adaptive traits (\emph{rufipogon} is adapted to forested wetland environments, whereas \emph{nivara} is adapted to dryer conditions and has life cycle adaptations to survive grazing pressure), suggesting that either offered (and offers) adaptive genetic potential.
%Domesticated varieties have likewise adapted to local environments.
%The \emph{japonicas} are generally more adapted to upland temperate regions (and are split into subgroups adapted to more temperate or more tropical conditions), the \emph{indicas} are best suited for tropical lowland environments, \emph{aus} matures quickly and tolerates drought stress, and two deepwater varieties (\emph{rayada} and \emph{ashwina}) are said to be selected for the environment along the Ganges river \cite{vaughan2008evolving, garris2005genetic}.
%Crop-wild hybridization is frequent enough that nuisance weedy hybrids commonly arise amongst cultivated populations and must be managed with difficulty \cite{vaughan2008evolving}.
%However, as with many other domesticated crops, gene flow between wild and domesticated rices is highly asymmetric (estimates of wild rice admixture in domesticated rice are less than 5 percent \cite{wang2017asian}).
%This asymmetry is due in part to the closed floret architecture of the domesticated rice, which hinders outcrossing.
%Still, barriers to crop-to-wild introgression may have been less severe during early domestication as the inbreeding reproductive system of domesticated rice would not have been as firmly established \cite{vaughan2008evolving}.
%Furthermore, the contemporary distribution of wild rice does not capture the full range and diversity of wild rice during early domestication and range expansion of Asian rice.

>>>>

The relationship between maize (\emph{Zea mays} ssp. \emph{mays}) and the teosinte \emph{Zea mays} ssp. \emph{mexicana} (hereafter, \emph{mexicana}) offers a prime case study of adaptive wild-to-crop introgression.
Maize was domesticated from \emph{Zea mays} ssp. \emph{parviglumis} (hereafter, \emph{parviglumis})  approximately 9,000 BP in the lowlands of the Balsas River Valley in Mexico \cite{matsuoka2002single}.
From this domestication center, maize spread into the highlands of the Mexican Central Plateau, where it came into sympatry with \emph{mexicana}.
Introgression from \emph{mexicana} to maize in the Central Plateau has been reported based on both morphological \cite {wilkes1977} and molecular \cite{vanHeerwaarden2011, doebley1987} data.
However, Hufford et al. \cite{Hufford2013} first localized \emph{mexicana} introgression to chromosomal regions and provided evidence that it was likely adaptive.
The authors identified nine genomic regions in several maize populations which consistently showed evidence of \emph{mexicana} introgression based on chromosome painting methods including HAPMIX (Figure \ref{fig:introgressionMaize}).
These introgressed segments overlapped with QTL that had previously been found to control anthocyanin content and leaf macrohairs \cite{lauter2004}, traits known to be adaptive at high elevation.
In a growth chamber experiment, the authors demonstrated that maize populations with \emph{mexicana} introgression showed greater plant height (a proxy for fitness) under highland environmental conditions than populations that lacked introgression.
Height differences were not detected under lowland conditions.

Populations of \emph{mexicana} cannot be found outside of the highlands of Mexico, yet maize has colonized and adapted to high elevation in a number of other regions.
A recent study \cite{wang2017} employed the ABBA-BABA and $\hat{f_{d}}$ statistics to evaluate whether maize with \emph{mexicana} introgression was transferred to other highland regions or whether highland adaptation was obtained \emph{de novo} outside of Mexico.
Overall, analyses revealed that  \emph{mexicana} introgressions were transferred to Mesoamerican high elevation regions in Guatemala and the southwestern United States, but more distant high elevation regions (\emph{e.g.,} the Andes) showed no \emph{mexicana} ancestry.

%Since \emph{mexicana} haplotypes appear to have played an important role historically in adaptation of maize to challenging high-elevation conditions, modern breeding programs may further benefit from drawing on \emph{mexicana} germplasm, particularly in regions like the Andes where \emph{mexicana} alleles are not known to have spread.

\begin{figure}[b]
	\centering
	\includegraphics[width=17.35cm]{./FigsAndTables/Figure1new}
	\caption{Hapmix vectors indicating introgression level from \emph{mexicana} to Mexican highland maize on chromosome 4. Data adopted from Hufford et al. \cite{Hufford2013}. ELP: EL Porvenir; IXT: Ixtlan ; NAB: Nabogame; OPO: Opopeo; PUR: Puruandiro; SAC: Santa Clara; SAP: San Pedro; TEN: Tenango del Aire; XOC: Xochimilco. Insets showed the phenotypes of macrohairs on lowland (red) and highland (blue) maize stems. The black bar indicated one of the QTLs for macrohairs and pigment density in Lauter et al. \cite{lauter2004}.}
	\label{fig:introgressionMaize}
\end{figure}

@@

Barley (\emph{Hordeum vulgare} subsp. \emph{vulgare}) was domesticated at least twice roughly 8,000 to 10,000 BP: once from the wild subsp. \emph{spontaneum} in the Fertile Crescent and once from subsp. \emph{spontaneum} var. \emph{agriocrithon} in Tibet \cite{takahashi1955origin, badr2000origin, azhaguvel2007phylogenetic, haberer2015barley, ren2013tibet, dai2012tibet}.
Presently, the distribution of subsp. \emph{spontaneum} stretches from the eastern Mediterranean through the Middle-East to west-central Asia, spanning clines in temperature, precipitation, soil type, and altitude \cite{nevo2010drought}.
Barley-\emph{spontaneum} hybrids are fertile and commonly found when these taxa co-occur.
Signatures of contemporary gene flow between wild and domesticated barley have been detected across distances greater than a kilometer \cite{hillman2001new}.

Poets and co-authors \cite{Poets2015} recently investigated the range-wide contribution of wild barley to landraces, assessing both genome-wide and geographical patterns.
This study identified several lines of evidence consistent with wild introgression aiding the dispersal and adaptation of domesticated barley.
Genomic regions of shared ancestry were detected linking particular landraces to numerous wild relative populations, suggesting landraces may have received wild introgression on a continual basis during post-domestication expansion.
However, barley landraces showed an excess of ancestry from nearby wild relatives, indicating a prevalence of local and potentially adaptive gene flow.
Limited linkage disequilibrium and small tracts of identity by state indicated even locally introgressed chromosomal regions are old, perhaps dating to the early expansion of barley  post-domestication.
While these results are suggestive, wild barley haplotypes have yet to be definitively linked to specific local adaptations in landraces.

@@

The details of Asian rice (\emph{Oryza sativa}) domestication are still debated.
Certain genetic and archaeobotanical evidence point toward independent domestications of the two prominent varietal groups \emph{japonica} and \emph{indica} from the wild species \emph{Oryza rufipogon} (\emph{rufipogon} hereafter) in China and the Indian Ganges plain, respectively \cite{fuller2010consilience}, with a potential third domestication event giving rise to \emph{aus} in Bangladesh or central India \cite{civavn2015three}.
Other studies support a single domestication in China, with later divergence of \emph{japonica} and \emph{indica} \cite{molina2011molecular, Huang2012} during crop exapnsion.
For example, Huang and colleagues \cite{Huang2012} developed a genetic map of rice variation, which they used to measure genetic distance between a range-wide sample of wild and domesticated rice, finding \emph{japonica} was domesticated near the Pearl River in Guangxi province, China, and that \emph{indica} was likely the result of hybridization between \emph{japonica} and local \emph{rufipogon} populations in Southern and South-eastern Asia.
In a re-examination of these same data, \cite{civavn2015three} found evidence supporting independent domestications of \emph{japonica}, \emph{indica}, and \emph{aus}, as well as a hybrid origin (\emph{japonica} x \emph{aus}) of \emph{aromatic} rice.
A recent third analysis by Choi and co-authors \cite{choi2018multiple} compared these two disparate results and the methods utilized therein, concluding that domestication alleles (including \emph{LABA1}, \emph{PROG1}, and \emph{sh1}) arose during a single domestication event of \emph{japonica}, and that these alleles introgressed into several wild \emph{rufipogon} subpopulations which thereby became the progenitors of other Asian rice varieties.

The findings of \cite{choi2018multiple} bear similarities to a hypothesis posited by Vaughan and colleagues \cite{vaughan2008evolving}.
According to this hypothesis, domestication alleles arose in a single cultivated rice population and subsequently introgressed into diverse cultivated populations (some \emph{japonica}-like, some \emph{indica}-like).
As these now fully domesticated populations spread further into new environments, they potentially received introgression from locally-adapted wild relatives, retaining alleles that improved fitness.
While the precise history of domesticated rice remains in question, multiple lines of evidence indicate diverse wild populations have contributed to domesticated germplasm and suggest adaptive introgression may have played a role during the expansion of this important crop.
